|

### 2021-11-13 07-Chp-Triangulation.Rmd
End of day: 40371

<!----------------------------------------------------------------------------->
&nbsp;

@@

It has been widely used in the literature, and is comparable to Bradford-Hill's criteria of "consistency", that is the replication of an observed relationship across several different contexts.[@hill1965] In this framework,

@@

### Terminology

@@

@@

Previous examples of this approach have used criteria such as

In contrast, _quantitative triangulation_ attempts to obtain a overall numerical result by synthesising the different types of evidence.

However, previous attempts at triangulation ignore the potential for varying and opposing sources of bias within a evidence source. For example,

The terminology used in describing different mechanisms by which an observed result may  complicate interpretation. This section clearly defines how each are used in h

Here, internal bias is used in terms of

In contrast, external bias (also called indirectness, or )

Additionally, potential biases can be additive or proportional. Pro

Talk about problems of sclae - assuming bias direction at the study level works well if you only have one study per design category as in the example by Ference et al (e.g. IPD analysis of )

@@

1. Define the causal question of interest
1. Identify of relevant evidence sources
1. Assess(and visualise) the extent and direction of bias in each result
1. Define modifying terms for bias and indirectness in each result
1. Perform bias-adjusted meta-analysis using bias-/indirectness-adjusted results

@@

As indicated in the options above, the response t this prompt is used both to define the direction and type (additive or proportional) of bias.

@@

For example, for a "High" additive bias in $j$th domain of the $i$th study is deifined as $\delta_{i,j}^{\mathrm{High}}$ and uncertainty around this estimate is given by a distribution:

$$
\delta_{i,j}^{\mathrm{High}} = f(\mu_{i,j}^{\mathrm{High}}, \sigma_{i,j}^{\mathrm{High}})
$$

@@

For the purposes of this example analysis,

High
Some concerns

These figures were sense-checked against

However, this also allows us to a priori define common modifying values across all domains of bias. In a bias-adjusted meta-analysis using this approach, a high risk of bias in the will have the same modifying effect on the results.

As such, a simulation approach was taken to explore the impacts of several different strategies

@@

####	Assign modifying values to indirectness

As I am attempting to identify the causal effect across different study designs, I needed to . This is not necessary if just summarising the literature.

@@

In the original method, this was assessed against 4 categories of internal bias.

@@

"One important difference between these approaches that will need to be addressed in the triangulation is that RCTs relate to short-term modification of lipid levels, whereas Mendelian randomisation studies typically refer to lifelong exposures; and traditional observational epidemiology approaches may refer to short or longterm exposure (although are often not explicit about this). The triangulation will take these into account, for example by superimposing the studies on a hypothetical model for exposure-outcome relationship across the life course."

<!-- From NIHR applications -->

<!-- "If the aim of a meta-analysis is to summarize the literature rather than to address a target question, there is no need to adjust for external biases since differences between the (idealized) studies are accepted as diversity. The need to adjust for internal biases remains and our method can be applied directly to these." Taken from https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2667303/ - good explanation on difference between qualitative and quantitative triangulation. -->

@@

Once the total internal and external additive and proportional biases for each result were computed, they were used to define an adjusted estimate for each result included in the synthesis

Using the method defined in Turner _et al_,[@turner2009] the adjusted estimate for each result is defined as:

$$
\hat{\theta} = \frac{y_i - \mu_{i\delta}^{\mathrm{I}} - \mu_{i\beta}^{\mathrm{E}}\mu_{i\delta}^{\mathrm{E}}}{\mu_{i\beta}^{\mathrm{I}}\mu_{i\delta}^{\mathrm{E}}}
$$

where $\mu_{i\delta}$ denotes total additive and $\mu_{i\beta}$ the total proportional bias, and the $^{\mathrm{I}}$ and $^{\mathrm{E}}$ superscripts denote internal and external (indirectness) biases respectively.

using the same terms, the standard error of this estimate, $\hat{\theta}$, is then calculated as:

$$
\mathrm{SE}(\hat{\theta})=\left(\frac{1}{\mu_{i \beta}^{\mathrm{I}} \mu_{i \beta}^{\mathrm{E}}}\right)^{2}\left(s_{i}^{2}+\left(\mu_{i \beta}^{\mathrm{I}} 2+\sigma_{i \beta}^{\mathrm{I}}\right)\left(\hat{\theta}^{2} \sigma_{i \beta}^{\mathrm{E} 2}+\sigma_{i \delta}^{\mathrm{E} 2}\right)+\sigma_{i \beta}^{\mathrm{I}}\left(\hat{\theta} \mu_{i \beta}^{\mathrm{E}}+\mu_{i \delta}^{\mathrm{E}}\right)^{2}+\sigma_{i \delta}^{\mathrm{I} 2}\right)
$$

@@

#### Software

@@

@@

@@

@@

Main takeway is that heteorgeneity between study designs is substantially reduced

@@

This chapter has demonstrated how new developments in systematic risk of bias assessment.

Future quantitative triangulation should move beyond the concept of comparing results at the approach level, and instead focus on the inherent threats to internal validity present in each specific result relevant to the causal question.

While this makes the process both more labour intensive and complex, to assume that, for example, all retrospective studies are equally at risk is to overlook. While ti could be true that all retrospective studies share some _minimal level_ of bias, there are several opportunities. Failing to separate these the individual results loses information, particularly given the now widespread availability of tools to systematically assess the biases present in each result.

In this chapter, I have presented a new method of visualising both the level and expected direction of bias inherent to each result, and suggested how this could be incorporated with a bias-adjusted meta-analysis approach to work towards a generalised qunatitative triangulation framework.

I've highlighted the limitations of this method, as the correct distributions of modifying values in each risk-of-bias domain should be driven by meta-epidemiological studies which do not yet exist, framing them as opportunities for future research.

@@

For example, if it was . New software for performing risk of bias assessments, being developed by the Bristol Appraisal and review of Research group has this as a secondary methodological aim.

@@

@@

#### Defining appropriate levels of bias

Previous work on assessing the impact of different sources of bias on effect estimates . For example, a study examining the effect of found that studies with

Previous work examining

@@

Ideally, the modifying values

@@

One potential - however, the estimation of an appropriate adjustment value for these studies would be substantially more challenging.

@@

#### Type of bias (proportional vs additive) not shown on figure

@@

### Alternative approaches

This method is also different from previous implementations of the bias-adjusted meta-analysis approach, in that for each domain for each study, the bias can be either proportional or. IN previous implementations, distributions for each source of bias (pre-domain-based tools) were elicted on both the additive and proportional scales.[@turner2009] Puts more pressure on the assessor to correct sum additive and proportional biases highlighted, and answer with only a single option.

#### E-values

Present as an alternative to putting priors on the level of bias due to confounding.

E-values work the other way around - get result then work out how strong a biasing factor would have to be to negate the result. IN other words, they represent an estimate ho how strong the total residual confounding would have to be in order to full explain away the observed result.[@vanderweele2017]

One major problem is that they can only be used to evaluate negating biases, not other biases.

How strong unmeasured confounding would have to be in order to explain the observed effect.

Heavily criticised in a range of papers.

Also, no useful to focus on a specific bias in the context of risk of bias domains

As an example, I calculated the E-value required to attenuate the hamrful association observed between statin use and vascular dementia in the CPRD analysis ( `r hold()`` ) presented in Chapter \@ref(cprd-heading). Assuming an outcome prevalence of <15%, an E-value of 3.31 was identified

This calculation was performed using the XXX R package.

<!-- TODO Ensure that the example is provided first, as you then go on to disparage it - in the example given, the main driver of bias is likely to be confounding (though some may be due to misclassification) -->

However, this approach is only useful if the only source of bias in a result is due to uncontrolled confounding (as assessed by a domain-based risk of bias tool).

Otherwise, it's interpretation is more challenging, as the biasing effect is incorrectly assinged complete to confoudning, thus overestimating the expected effect of confounding.

the contribution of confounding to the bias in a result, and while it may be applicable if the only bias assessed to

@@

As the need and appetite for snthesis of different, future iterations of the risk of bias tools should take this into account. Simple things like harmonising the risk of bias levels across tools would be a good first move towards a more integrate suite of tools. Similarly, software that implements the risk-of-bias tools should both allow for direction of bias question and carefully consider how this data will be exported.

@@

While the limitations inherent to this method are important, and limit the , they reflect a more detailed appreciation of the complexities of risk of bias assessment and triangulation as whole (i.e. not applying expected directions of effects to a ).

This approach also builds on existing approaches employed previously by "exploding out" the results of a systematic review/meta-analysis and considering the effects of bias in each individual result seperately. This more granular approach is to be preferred over assigning a bias judgement and direction to the summary effect produced by a meta-analysis,  which may mask the different biases applicable to each unique result.

@@

Triangulation as a extension of evidence synthesis.

@@

@@

>>>>

@@

It has been widely used in the literature, and is comparable to Bradford-Hill's criteria of "consistency", that is the replication of an observed relationship across several different contexts.[@hill1965] A good working definition of the triangulation is:

> The practice of strengthening causal inferences by integrating results from several different approaches, where each approach has different (and assumed to be largely unrelated) key sources of potential bias.[@lawlor2016]

This represents a significant step forward from the current evidence synthesis practice of synthesising evidence from only one approach (e.g. RCT, observational study, etc.).

However, previous attempts at triangulation have ignored the potential for varying and opposing sources of bias within an approach, assigning a level and direction of bias to entire categories of studies.

Recent development in bias assessment have a allowed fro a much more granular consideration of the threats to the internal validity of a result.

For example, while it may be true that all non-randomised studies share some _minimal level_ of bias, investigation of the biases applicable to each individual result gives useful additional information, and is essential for any attempt at quantitative triangulation.

This chapter, in addition to providing a narrative comparison of the existing and new evidence identified in this thesis, attempts to combine the results numerically by building on existing work on this topic.

In the following section, I define important terms used throughout this chapter.

@@

### Terminology {#tri-terminology}

<!-- Discuss and defined three things: -->

<!-- * Qualitative vs quantitative triangulation -->
<!-- * Bias vs. indirectness -->
<!-- * Additive vs proportion effect of bias/indirectness -->

@@

Triangulation exercises can be characterised as either qualitative or quantitative.

@@

In contrast, _quantitative triangulation_ attempts to obtain a overall numerical result by synthesising the different types of evidence. This method is in its infancy, given important methodological challenges that are illustrated in this Chapter.

Both

Bias (or internal bias) in the result is defined as systematic deviation from result the "idealised" version of the study (that is, )

In contrast, the indirectness of the result (also called generalisability or applicability) is defined here as the discrepancy between the research question addressed by the "idealised" study and the causal question of interest. Problems with indirectness can be arise from differences in population, intervention/exposure, outcomes and

Of note, the original method employed in this chapter refers indirectness as "external bias" - however. This distinction is important, as bias and indirectness describe two different issues, and referring to both as bias

In addition, indirectness is the widely used term, including by widely-adopted methods such as the GRADE framework.[@guyatt2011]

__Additive vs proportional effects__

The effects of both bias and indirectness can be either additive or propotional.

Additive bias/indirectness is

Both bias and indirectness can be defined as additive or propotional. Additive bias/indirectness is that which can shift the p

@@

1. Define the causal question of interest
1. Identify of relevant evidence sources
1. Assess(and visualise) the extent and direction of bias in each result
1. Define modifying terms for bias and indirectness in each result
1. Perform bias-adjusted meta-analysis using bias-adjusted results

@@

As indicated in the options above, the response t this prompt is used both to define the direction and type (additive or proportional) of bias (see ).

@@

For example, for a "High" additive bias in $j$th domain of the $i$th study is deifined as $\delta_{i,j}^{\mathrm{High}}$ and uncertainty around this estimate is given by a distribution

\begin{equation}
  \delta_{i,j}^{\mathrm{High}} = f(\mu_{i,j}^{\mathrm{High}}, \sigma_{i,j}^{\mathrm{High}})
  (\#eq:name)
\end{equation}

@@

@@

<!-- TODO In the absence of any empirical data regarding the priors, I used the data from the expert elicication approach to calculate both the mean and max. These estimates, in addition to conversations with topic experts -->

For the purposes of this illustrative analysis and to highlight the generalised nature of the code used to perform the analysis (see Section \@ref(tri-software)), I defined two sets of modifying distributions (Table 1). In the first scenario, the relationship between the position assigned to each judgement is linear. In the second scenario, . In both cases, the variance of the "High" risk of bias judgement was defined to be great than the "Moderate" risk of bias judgement. The two scenarios are illustrated in Figure \@ref(fig).

Reasonable values for the position and spread of the distributions were informed by calculating the mean and maximum adjustment values as defined by the original elicitation exercise (see Appendix \@ref(?) for more details on how these were calculated). [@turner2009] <!-- TODO Link this to open data! -->

Where the direction of bias in a domain was unpredictable for a given result, the mean of the adjustment distribution for that domain, but the appropriate variance for (i.e. no predicted direction, but increased uncretainty in line with judgement).

####	Assessing and assigning modifying values to indirectness

@@

In the original method, this was assessed against 4 categories of external bias (population/intervention/control/outcome)

@@

A key example here is that

All indirectness bias were defined as being proportional. As described i

This is also in line with previous efforts to assess the adjustment needed due to indirectness

As crude example focusing on the intervention , versus the causal question of the effect of LDL-c in mid-life on Alzheimer's disease, studies of lipids levels in this time period would require minial adjustment, whereas studies
...

@@

Once the total additive and proportional biases for each result were computed, they were used to define an adjusted estimate for each result included in the synthesis.

Using the method defined in Turner _et al_,[@turner2009] the adjusted estimate for each result, $\hat\theta$, is defined as:

\begin{equation}
  \hat{\theta} = \frac{y_i - \mu_{i\beta}^{\mathrm{In}}\mu_{i\delta}^{\mathrm{In}} - \mu_{i\delta}^{\mathrm{B}}}{\mu_{i\beta}^{\mathrm{B}}\mu_{i\beta}^{\mathrm{In}}}
  (\#eq:adjusted-mean)
\end{equation}

where $\mu_{i\delta}$ denotes total additive and $\mu_{i\beta}$ the total proportional adjustment, and the $^{\mathrm{B}}$ and $^{\mathrm{In}}$ superscripts denote wether the adjustment relates to internal bias or indirectness respectively.

Using the same terms, the standard error of this estimate, $\hat{\theta}$, is then calculated as:

\begin{equation}
  \mathrm{SE}(\hat{\theta})=\left(\frac{1}{\mu_{i \beta}^{\mathrm{B}} \mu_{i \beta}^{\mathrm{In}}}\right)^{2}\left(s_{i}^{2}+\left(\mu_{i \beta}^{\mathrm{B}} 2+\sigma_{i \beta}^{\mathrm{B}}\right)\left(\hat{\theta}^{2} \sigma_{i \beta}^{\mathrm{In} 2}+\sigma_{i \delta}^{\mathrm{In} 2}\right)+\sigma_{i \beta}^{\mathrm{B}}\left(\hat{\theta} \mu_{i \beta}^{\mathrm{In}}+\mu_{i \delta}^{\mathrm{In}}\right)^{2}+\sigma_{i \delta}^{\mathrm{B} 2}\right)
  (\#eq:adjusted-se)
\end{equation}

@@

#### Software {#tri-software}

@@

Show result and bias

Reproduce version of Table 2 rom the "proposed method" paper for indirectness

Calculate internal-bias adjusted

Calculate both-adjusted

@@

Midlife LDL-c on Alzheimer's disease

@@

<!----------------------------------------------------------------------------->
(ref:ldlDemBiasDirection-cap) __Effect of LDL-c on All-cause dementia in late life:__

(ref:ldlDemBiasDirection-scap) Effect of LDL-c on All-cause dementia in late life

```{r ldlDemBiasDirection, echo = FALSE, results="asis", fig.pos = "H", fig.cap='(ref:ldlDemBiasDirection-cap)', out.width='100%', fig.scap='(ref:ldlDemBiasDirection-scap)'}
knitr::include_graphics(file.path("figures/misc/UoBcrest.pdf"))
```
<!----------------------------------------------------------------------------->

@@

Main takeaway is that heterogeneity between study designs is substantially reduced

@@

Thsi chapter has narratiely synthesised the exsiting and new evidence from the previous chapters, highlighting the lim

In addition, it has sought to provide a generalised framework for quantitative triangulation, building on existing methods for systematic domain-based risk of bias assessment and bias-/indirectness-adjusted meta-analysis.

Future quantitative triangulation should move beyond the concept of comparing results at the approach level, and instead focus on the inherent threats to internal validity present in each specific result relevant to the causal question. While this makes the process both more labour intensive and complex, to assume that, for example, all retrospective studies are equally at risk is to overlook.

@@

For example, if it was . New software for performing risk of bias assessments, being developed by the Bristol Appraisal and Review of Research group has this as a secondary methodological aim.

@@

Appreciation that both the risk of bias and indirectness of a result should be assessed is growing. Indeed, some existing risk of bias for the assessment of diagnostic test accurary[@whiting2011a] and prediction models[@moons2019] consider indirectness (termed applicability in these tools) alongside the assessment of bias. <!-- TODO Cite PROBAS and QUADAS -->

@@

#### Defining appropriate values for bias/indirectness adjustment

Much of the attention to date has been on assessing the impact of different sources of bias on effect estimates has focused predominantly on randomised controlled trials.

For example, a study examining the effect of found that only a subset of domains appear to impact the result, and that this effect is modest (10% change).[@savovic2018] However, this work examined RCTs, and these findings may not hold for other non-randomised study designs. <!-- TODO Double check this citation actually says what you think it does! -->

@@

@@

As illustrated in Table \@ref(tab:robLevelsMapping-table), studies at critical risk of bias were not included in the analysis.

one potential limitation of the  - however, the estimation of an appropriate adjustment value for these studies would be substantially more challenging.

@@

#### Type of bias (proportional vs additive) not shown on bias direction plot

@@

### Related approaches

<!-- TODO Need more here on the other approaches discussed in the recent comparison paper -->

@@

As the need and appetite for synthesis of different, future iterations of the risk of bias tools should take this into account. Simple steps like harmonising the risk of bias levels across tools would be a good first move towards a more integrate suite of tools. Similarly, software that implements the risk-of-bias tools should both allow for direction of bias question and carefully consider how this data will be exported.

@@

While the limitations inherent to this method are important, and limit the , they reflect a more detailed appreciation of the complexities of risk of bias assessment and triangulation as whole (i.e. not applying expected directions of effects to an entire category of studies ).

This approach also builds on existing approaches employed previously by "exploding out" the results of a systematic review/meta-analysis and considering the effects of bias in each individual result separately. This more granular approach is to be preferred over assigning a bias judgement and direction to the summary effect produced by a meta-analysis,  which may mask the different biases applicable to each unique result.

@@

Triangulation, both qualitative and quantitative, should be considered an extension of evidence synthesis, and so. Triangulation exercises should be based on comprehensive systematic reviews to create. If needed, evidence gaps can be addressed through additional studies, as in the case of this analysis,

### Outputs

Software

Annotated example dataset

this ties in with the open data sharing of the by-assessor elictation data from, which i sued in this analysis to frame the modifying values assigned to

Generalised version of code, allowing for different distributions in each domain of bias (which will be handy once we get a better idea of what they should be).

@@

In this chapter, I have presented a new method of visualising both the level and expected direction of bias inherent to each result, and suggested how this could be incorporated with a bias-adjusted meta-analysis approach to work towards a generalised quantitative triangulation framework.

Finally, I've highlighted the limitations of this method, as the correct distributions of modifying values in each risk-of-bias domain should be driven by meta-epidemiological studies which do not yet exist, framing them as opportunities for future research.

@@

The implications of this work to clinical practice, public health and future research is considered in the following chapter.

### 2021-11-13 08-Discussion.Rmd
End of day: 40371

@@

It has analysed two previously unalysed datasets, adding useful information to the evidence base

@@

The clinical recommendations that can be drawn from the work in this thesis are limited, due to the

however, a key limiting factor is the absence of a detailed pathological mechanism for Alzheimer's disease. Additionally, vascular dementia

@@

@@

>>>>

Given the detailed discussion

## New evidence

Found slight raised risk of dementia in CPRD patients () with higher LDL-c measured at midlife, driven by the finding for Alzheimer's disease [@iwagami2021]. Not inconsistent with the findings presented in this thesis

## Lessons learned

There are

@@

It has analysed two previously unanalysed datasets, adding useful information to the evidence base.

@@

Across multiple sources of evidence, there was no consistent effect of blood lipids, or . Some

The clinical recommendations that can be drawn from the work in this thesis are limited, due to the low strength of evidence related to the

There was very we. Conversely however, there was an absence of evidence for a harmful effect of statins on. Given the reality that given the choice, given the conflicting evidence

Patients must live to a sufficient age to be at risk of dementia outcomes, which is substantially less like if high lipid levels in mid-life.

The role of lipid-lowering

As with many many studies of dementia outcomes, a key limiting factor of any analysis is the absence of a detailed pathological mechanism for Alzheimer's disease. It could well be that statins do have a true effect independently of lipid levels, though based on the certainty of evidence in NRSI, this is unlikely.

@@

For other dementia outcomes,

@@

Several avenues for future research, buiding on the work presented here,

###
